---
title: "An Introduction to goSorensen R-Package"
author:
- name: Jordi Ocaña
  email: jocana@ub.edu
  affiliation: Department of Genetics, Microbiology and Statistics, Statistics Section, University of Barcelona
- name: Pablo Flores
  email: p_flores@espoch.edu.ec
  affiliation: Escuela Superior Politécnica de Chimborazo (ESPOCH), Facultad de Ciencias, Carrera de Estadística.
package: goSorensen
abstract: >
  This vignette provides an introduction to **goSorensen**  R-package. The goal of goSorensen is to implement an inferential statistical method to compare gene lists (proving equivalence) in terms of their biological significance as expressed in the Gene Ontology GO. This method is introduced in the paper "An equivalence test between features lists, based on the Sorensen - Dice index and the joint frequencies of GO node enrichment", available in <https://rdcu.be/cOISz>.
  
  
  For a given ontology (BP, CC, or MF) and GO-Level, both enriched and non-enriched GO terms are identified for the compared lists. This enrichment can be summarized in contingency tables for each pair of compared lists. The dissimilarity between gene lists is assessed using a measure based on the Sorensen-Dice index. The fundamental guide principle is that two gene lists are considered similar if they share a significant proportion of common enriched GO terms.
  
output:
  BiocStyle::html_document:
    toc_float: false
    toc: false
vignette: >
  %\VignetteIndexEntry{An introduction to equivalence test between feature lists using goSorensen.}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Annotation, GO, GeneSetEnrichment, Software, Microarray, Pathways, GeneExpression, MultipleComparison}
  %\VignetteEncoding{UTF-8}
  
---


```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r css, echo=FALSE, results='asis'}
cat("
<style>
/* Ocultar inicialmente el contenido colapsable */
.collapsible-content {
    display: none;
}

/* Cambiar el puntero del ratón en los encabezados colapsables */
.collapsible-header {
    cursor: pointer;
}

.collapsible-header:hover {
    text-decoration: underline;
}

/* Título de Nivel 1 */
h1 {
    font-size: 1.2em; 
    color: #0e5775; 
    font-weight: bold;
}

/* Título de Nivel 2 */
h2.collapsible-header {
    font-size: 1.1em; 
    color: #12769f; 
    margin-left: 20px; 
    font-weight: bold;
}

/* Título de Nivel 3 */
h3.collapsible-header {
    font-size: 1.0em; 
    color: #16a3dc; 
    margin-left: 40px; 
    font-weight: bold;
}

/* Mantén el contenido alineado con los encabezados */
.collapsible-content {
    margin-left: inherit;
}

/* Ajuste opcional: evitar modificar el tamaño o color del encabezado */
</style>
")
```

```{r js, echo=FALSE, results='asis'}
cat("
<script>
document.addEventListener('DOMContentLoaded', function () {
    const excludedSections = ['Abstract', 'Package', 'Author Information', 'Date'];

    // Busca los encabezados principales
    const headers = document.querySelectorAll('h2, h3');
    headers.forEach(header => {
        const headerText = header.textContent.trim();

        // Excluir encabezados específicos
        if (excludedSections.some(section => headerText.includes(section))) {
            return;
        }

        // Encuentra todo el contenido hasta el próximo encabezado
        const content = [];
        let next = header.nextElementSibling;
        while (next && !/^H[1-6]$/.test(next.tagName)) {
            content.push(next);
            next = next.nextElementSibling;
        }

        // Crear contenedor colapsable
        const contentWrapper = document.createElement('div');
        contentWrapper.className = 'collapsible-content';
        content.forEach(node => contentWrapper.appendChild(node));

        // Agregar funcionalidad de colapsar/expandir
        header.classList.add('collapsible-header');
        header.addEventListener('click', () => {
            const isVisible = contentWrapper.style.display === 'block';
            contentWrapper.style.display = isVisible ? 'none' : 'block';
        });

        // Insertar el contenido después del encabezado
        header.parentNode.insertBefore(contentWrapper, next);
    });
});
</script>
")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = ""
)
```

```{r env, message = FALSE, warning = FALSE, echo = TRUE}
library(goSorensen)
```

# PRELIMINARIES. 
## Theoretical Framework of Reference. 
Given two gene lists, $L_1$ and $L_2$, (the data) and a given set of $n$ Gene Ontology (GO) terms (the frame of reference for biological significance in these lists), goSorensen makes the necessary compute devoted to answering the following question: Is the dissimilarity between the biological information in both lists negligible? In other words, are both lists functionally equivalent?

We employ the following metric derived from the Sorensen-Dice index to quantify this dissimilarity:

\begin{equation*}
\hat{d_S} = 1 - \dfrac{2n_{11}}{2n_{11} + n_{10} + n_{01}} \\
\hat{d_S} = 1 - \dfrac{\frac{2n_{11}}{n}}{\frac{2n_{11}}{n} + \frac{n_{10}}{n} + \frac{n_{01}}{n}} \\
\hat{d_S} = 1 - \dfrac{2\widehat{p}_{11}}{2\widehat{p}_{11} + \widehat{p}_{10} + \widehat{p}_{01}}
\end{equation*}

where:

- The joint enrichment is the number of enriched GO terms in both lists. It is represented by $n_{11}$ in absolute frequency or $\hat{p}_{11}$ in relative frequency. 
- The marginal enrichment refers to the number of enriched GO terms found in only one list, but not in the other. It is represented by $n_{10}$ and $n_{01}$ in absolute frequencies or $\hat{p}_{10}$ and $\hat{p}_{01}$ in relative frequencies. For example, $n_{10}$ represents the number of GO terms enriched in list $L_1$ but not in list $L_2$.
- The non enrichment is the number of GO terms non enriched in any lists. It is represented by $n_{00}$ in absolute frequency or $\hat{p}_{00}$ in relative frequency.



The enrichment frequency can be represented in a $2 \times 2$ contingency table, as follows:

|                           	| **enriched in $L_2$** 	| **non-enriched in $L_2$** 	|          	|
|:-------------------------:	|:---------------------:	|:-------------------------:	|:--------:	|
|   **enriched in $L_1$**   	|        $n_{11}$       	|          $n_{10}$         	| $n_{1.}$ 	|
| **non-enriched in $L_1$** 	|        $n_{01}$       	|          $n_{00}$         	| $n_{0.}$ 	|
|                           	|        $n_{.1}$       	|          $n_{.0}$         	|    $n$   	|

We have demonstrated that $d_S$ asymptotically follows a normal distribution. In cases of low joint enrichment, a sampling distribution derived from the bootstrap approach demonstrates a better fit and provides more suitable results.

Given the following equivalence hypothesis test: 

\begin{equation*}
H_0:d_S \ge d_0 \\
H_1: d_S < d_0
\end{equation*}

Where $d_S$ represents the "true" Sorensen dissimilarity. If this theoretical measure is equivalent to zero, it implies that the lists compared $L_1$ and $L_2$ share a significant proportion of enriched terms GO, which can be interpreted as biological similarity. Equivalence is understood as an equality, except for negligible deviations, which is defined by the limit $d_0$

A not-so-arbitrary irrelevance limit is $d_0 = 0.4444$, or more restrictive $d_0=0.2857$

For the moment, the reference set of GO terms can be only all those GO terms in a given level of one GO ontology, either Biological Process (BP), Cellular Component (CC) or Molecular Function (MF).

For more details, see the reference paper.


## goSorensen Installation. 

goSorensen package must be installed with a working R version (\>=4.4.0). Installation could take a few minutes on a regular desktop or laptop. Package can be installed from Bioconductor, then it needs to be loaded using `library(goSorensen)`:

```{r, eval=FALSE}
if (!requireNamespace("goSorensen", quietly = TRUE)) {
    BiocManager::install("goSorensen")
}
library(goSorensen)
```

## Data Used in this Vignette. 

The dataset used in this vignette, `allOncoGeneLists`, is based on the gene lists compiled at <http://www.bushmanlab.org/links/genelists>, a comprehensive set of gene lists related to cancer. The package `goSorensen` loads this dataset using `data(allOncoGeneLists)`:

```{r}
data("allOncoGeneLists")
```

`allOncoGeneLists`  is an object of class list, containing seven character vectors with the ENTREZ gene identifiers of a gene list related to cancer.

```{r}
sapply(allOncoGeneLists, length)

# First 15 gene identifiers of gene lists atlas and sanger:
allOncoGeneLists[["atlas"]][1:15]
allOncoGeneLists[["sanger"]][1:15]
```

## Before Using goSorensen.  
Before using `goSorensen,` the users must have adequate knowledge of the species they intend to focus their analysis on. The genomic annotation packages available in Bioconductor provide all the essential information about many species. 


For the specific case of this vignette, given that the analysis will be done in the human species, the `org.Hs.eg.db` package must be previously installed and activated as follows:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
    BiocManager::install("org.Hs.eg.db")
}
```

```{r, message = FALSE, warning = FALSE, eval = TRUE}
library(org.Hs.eg.db)
```

Actually, the `org.Hs.eg.db` package is automatically installed as a dependency on `goSorensen`, making its installation unnecessary. However, for any other species, the user must install the correspondence genome annotation for the species to analyze, as indicated in the above code. 


In addition, it is necessary to have a vector containing the IDs of the universe of genes associated with the species under study. The genomic annotation package provides an easy way to obtain this universe. The ENTREZ identifiers of the gene universe for humans, necessary for this vignette, is obtained as follows:

```{r, message = FALSE, warning = FALSE, echo = TRUE}
humanEntrezIDs <- keys(org.Hs.eg.db, keytype = "ENTREZID")
```

In this same way, the identifiers of the gene universe can be obtained for any other species.

Other species available in **Bioconductor** may include:

- `org.Hs.eg.db`: Genome wide annotation for Humans.
- `org.At.tair.db`: Genome wide annotation for Arabidopsis
- `org.Ag.eg.db`: Genome wide annotation for Anopheles
- `org.Bt.eg.db`: Genome wide annotation for Bovine
- `org.Ce.eg.db`: Genome wide annotation for Worm
- `org.Cf.eg.db`: Genome wide annotation for Canine
- `org.Dm.eg.db`: Genome wide annotation for Fly
- `org.EcSakai.eg.db`: Genome wide annotation for E coli strain Sakai
- `org.EcK12.eg.db`: Genome wide annotation for E coli strain K12
- `org.Dr.eg.db`: Genome wide annotation for Zebrafish
- `org.Gg.eg.db`: Genome wide annotation for Chicken
- `org.Mm.eg.db`: Genome wide annotation for Mouse
- `org.Mmu.eg.db`: Genome wide annotation for Rhesus

Due to the extensive research conducted on the human species and the examples documented in `goSorensen` for this species, the installation of the `goSorensen` package automatically includes the annotation package `org.Hs.eg.db` as a dependency.

If you are working with other species, you must install the appropriate package to use the genomic annotation for those species."



# ENRICHMENT OF GO TERMS.
## For One List.
The first step is to determine whether the GO terms for a specific ontology and GO-Level are enriched or not enriched across the different lists to be compared. The `enrichedIn` function assigns `TRUE` when a GO term is enriched in a list and `FALSE` when it is not.

For example, for the list `atlas`, which is part of `allOncoGeneLists` and originates from a study on the human species, the enrichment of GO terms in the BP ontology at GO-Level 4 is determined as follows:

```{r, echo=FALSE}
options(max.print = 50)
```

```{r, eval=TRUE}
enrichedAtlas <- enrichedIn(allOncoGeneLists[["atlas"]],
           geneUniverse = humanEntrezIDs, 
           orgPackg = "org.Hs.eg.db",
           onto = "BP", GOLevel = 4)
enrichedAtlas
```

The result is a vector containing only the GO terms enriched (`TRUE`) by the `atlas` list in the BP ontology at GO-Level 4. 

The attribute `nTerms` indicates the total number of GO terms, both enriched (`TRUE`) and non-enriched (`FALSE`), by the list `atlas` in the BP ontology at GO-Level 4. To obtain this vector, the logit argument `onlyEnriched` (which is `TRUE` by default) must be set to `FALSE`, as follows:

```{r}
fullEnrichedAtlas <- enrichedIn(allOncoGeneLists[["atlas"]],
           geneUniverse = humanEntrezIDs, 
           orgPackg = "org.Hs.eg.db",
           onto = "BP", GOLevel = 4, 
           onlyEnriched = FALSE)
fullEnrichedAtlas
```

The full vector (`fullEnrichedAtlas`) is much larger than the vector containing only the enriched GO terms (`enrichedAtlas`), which implies a higher computational cost.

```{r}
# number of GO terms in enrichedAtlas
length(enrichedAtlas)

# number of GO terms fullEnrichedAtlas
length(fullEnrichedAtlas)
```

The length of `fullEnrichedAtlas` corresponds to the total number of GO terms in the BP ontology at GO-Level 4. In contrast, the length of `enrichedAtlas` represents only the number of GO terms that are enriched by the list `atlas`.


## For Two or More Lists.
For the seven lists of `allOncoGeneLists`, the matrix containing the GO terms enriched in at least one of the lists to be compared is calculated as follows:

```{r, echo=FALSE}
options(max.print = 100)
```

```{r, eval=FALSE}
enrichedInBP4 <- enrichedIn(allOncoGeneLists, 
                               geneUniverse = humanEntrezIDs, 
                               orgPackg = "org.Hs.eg.db", 
                               onto = "BP", GOLevel = 4) 
enrichedInBP4
```

```{r, echo=FALSE}
data("enrichedInBP4")
enrichedInBP4
```

To obtain the full matrix with all the GO terms in the BP ontology at GO-Level 4, we must set the argument `onlyEnriched`  (which is `TRUE` by default) to `FALSE`, as follows:

```{r, eval=FALSE}
fullEnrichedInBP4 <- enrichedIn(allOncoGeneLists, 
                               geneUniverse = humanEntrezIDs, 
                               orgPackg = "org.Hs.eg.db", 
                               onto = "BP", GOLevel = 4,
                               onlyEnriched = FALSE) 
fullEnrichedInBP4
```

```{r, echo=FALSE}
data("fullEnrichedInBP4")
fullEnrichedInBP4
```

The number of rows in the full matrix (`fullEnrichedInBP4`) is much larger than in the matrix containing only the GO terms enriched in at least one list (`enrichedInBP4`), which implies a higher computational cost.

```{r}
# number of GO terms (rows) in enrichedInBP4
nrow(enrichedInBP4)

# number of GO terms (rows) in fullEnrichedInBP4
nrow(fullEnrichedInBP4)
```

The number of rows in `fullEnrichedInBP4` corresponds to the total number of GO terms in the BP ontology at GO-Level 4. In contrast, the number of rows in `enrichedInBP4` represents only the GO terms enriched by at least one list from `allOncoGeneLists`, meaning that each row in this matrix contains at least one `TRUE.`

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the objects `enrichedInBP4` and `fullEnrichedInBP4`, which can be accessed using `data(enrichedInBP4)` and `data(fullEnrichedInBP4)`. 
Please consider these objects as illustrative examples only, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

The calculations illustrated in this vignette are based on the matrix containing GO terms enriched by at least one list (in our case, `enrichedInBP4`). In the illustrations provided for this vignette, there is no evidence to suggest that this matrix produces results different from the full matrix, which includes all GO terms for a specific ontology and level, including those that are not enriched in any of the lists being compared. This is very beneficial since the computational cost of processing is much lower than it could be.


# ENRICHMENT CONTINGENCY TABLES
## For a Specific Ontology and one GO-Level. 
### Contingency Tables for Two Lists.
The enrichment contingency tables produced by `goSorensen` are the direct result of performing cross-frequency tables between pairs of columns (lists) of the enrichment matrices determined in the Section 2.2 of this vignette.

For the specific case of two gene lists, the function `buildEnrichTable` computes the contingency table by accepting two vectors of the class `character` containing the IDs of the lists to be compared.
For instance, for the BP ontology at GO-Level 4, the contingency table representing the enrichment of GO terms in the lists `atlas` and `sanger` is obtained as follows:

```{r, eval=FALSE}
cont_atlas.sanger_BP4 <- buildEnrichTable(allOncoGeneLists$atlas, 
                                          allOncoGeneLists$sanger,
                                          listNames = c("atlas", "sanger"),
                                          geneUniverse = humanEntrezIDs, 
                                          orgPackg = "org.Hs.eg.db", 
                                          onto = "BP", 
                                          GOLevel = 4)
cont_atlas.sanger_BP4
```

```{r, echo=FALSE}
data("cont_atlas.sanger_BP4")
cont_atlas.sanger_BP4
```

The result is an enrichment contingency table of the class `table`.

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the object `cont_atlas.sanger_BP4`, which can be accessed using `data(cont_atlas.sanger_BP4)`.

Please consider this object only as illustrative example, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

### Contingency Tables for Two or More Lists.
Given $s$ ($s \geq 2$) lists to compare, the $s(s-1)/2$ possible enrichment contingency tables can also be obtained using the function `buildEnrichTable.` Instead of providing two vectors as the main argument, we provide an object of the class `list`, containing at least two elements, each of which contains the identifiers of the different lists to be compared.

For example, for the BP ontology at GO-Level 4, the $7(6)/2=21$ contingency tables calculated from the 7 gene lists contained in `allOncoGeneLists` are obtained as follows:
```{r, eval=FALSE}
cont_all_BP4 <- buildEnrichTable(allOncoGeneLists,
                                 geneUniverse = humanEntrezIDs, 
                                 orgPackg = "org.Hs.eg.db", 
                                 onto = "BP", 
                                 GOLevel = 4)
```

The result is an object of the class `tableList`, which is exclusive from `goSorensen` and contains all the possible enrichment contingency tables between the compared gene lists at GO-Level 4 for the ontology BP. Since the output is very large, it is not displayed it in this vignette.

An important attribute of this object is `enriched`, accessible via `attr(cont_all_BP4, "enriched")`, which contains the enrichment matrix obtained using the `enrichedIn` function. For this particular case, `attr(cont_all_BP4, "enriched")` contains exactly the same information as the object `enrichedInBP4` from Section 2.2 of this vignette.

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the object `cont_all_BP4`, which can be accessed using `data(cont_all_BP4)`.

Please consider this object only as illustrative example, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

## For More than One Ontology and GO-Level.
When you want to obtain contingency tables for two or more lists across multiple ontologies and more than one GO-Level, you can use the function `allBuildEnrichTable.`

For example to obtain the $7(6)/2=21$ contingency tables calculated from the 7 gene lists in `allOncoGeneLists` for the three ontologies (BP, CC, and MF) and for the GO-Levels from 3 to 10, you can use the function `allBuildEnrichTable` as follows:
```{r, eval=FALSE}
allContTabs <- allBuildEnrichTable(allOncoGeneLists,
                                   geneUniverse = humanEntrezIDs, 
                                   orgPackg = "org.Hs.eg.db", 
                                   ontos = c("BP", "CC", "MF"), 
                                   GOLevels = 3:10)
allContTabs
```


The result is an object of the class `allTableList`, which is exclusive from `goSorensen` and contains all possible enrichment contingency tables between the compared gene lists for the BP, CC, and MF ontologies, and for GO-Levels 3 to 10. Since the output is very large, it is not displayed in this vignette.

The attribute `enriched` is present in each element of this output, meaning that for each ontology and GO-Level contained in this object, there is an enrichment matrix similar to the one obtained with the function `enrichedIn`. For instance, by running the code `attr(allContTabs$BP$'level 4', 'enriched')`, you can access the enrichment matrix `enrichedInBP4` obtained in Section 2.2 of this vignette. 

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the object `allContTabs`, which can be accessed using `data(allContTabs)`.

Please consider this object only as illustrative example, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

# EQUIVALENCE TESTS.
## For a Specific Ontology and GO-Level.
### Equivalence Test for Two Lists.
The function `equivTestSorensen` allows you to compute the necessary outputs for performing an equivalence test to detect equivalence between gene lists.

For the specific case of two gene lists, you need to provide the function `equivTestSorensen` with two character vectors containing the IDs of the lists to be compared.

For example, using an asymptotic normal distribution, an irrelevance limit $d_0=0.4444$ and a significance level $\alpha=0.05$ (the default parameters), an equivalence test to compare the lists `atlas` and `sanger` for the BP ontology at GO-Level 4 can be performed as follows:
```{r, eval=FALSE}
eqTest_atlas.sanger_BP4 <- equivTestSorensen(allOncoGeneLists$atlas, 
                                             allOncoGeneLists$sanger,
                                             listNames = c("atlas", "sanger"),
                                             geneUniverse = humanEntrezIDs, 
                                             orgPackg = "org.Hs.eg.db", 
                                             onto = "BP", GOLevel = 4,
                                             d0 = 0.4444, 
                                             conf.level = 0.95)
eqTest_atlas.sanger_BP4
```

```{r, echo=FALSE}
data("eqTest_atlas.sanger_BP4")
eqTest_atlas.sanger_BP4
```

If the enrichment contingency table is available prior to performing the test, such as `cont_atlas.sanger_BP4` determined in Section 3.1.1, the execution time for the calculation is much shorter. To use `equivTestSorensen` with the contingency table as input, proceed as follows:
```{r}
equivTestSorensen(cont_atlas.sanger_BP4, 
                  d0 = 0.4444, 
                  conf.level = 0.95)
```
As you can see, both procedures produce the same result, but the last one (whenever possible) is much faster. Regardless of the procedure, the result is an object of class `equivSDhtest`, which is exclusive from `goSorensen`.

If you want to change the calculation parameters of the test, such as using a bootstrap distribution instead of a normal distribution, setting an irrelevance limit of $d_0 = 0.2857$ instead of $d_0 =0.4444$, or changing the significance level to $\alpha = 0.01$ instead of $\alpha = 0.05$, one option would be to use the `equivTestSorensen` function again with the new parameters. However, this would require performing all the calculations again, leading to additional computational costs, which increase as more tests are performed. Instead, the  function `upgrade` allows you to update the test output much more quickly by simply specifying the name of the object where the test results are stored and the new parameters you wish to apply, as shown below:

```{r}
upgrade(eqTest_atlas.sanger_BP4, d0 = 0.2857, 
        conf.level = 0.99, boot = TRUE)
```

### Equivalence Test for Two or More Lists.
Given $s$ ($s \geq 2$) lists to compare, the $s(s-1)/2$ possible equivalence tests can also be obtained using the function `equivTestSorensen` Instead of providing two vectors as the main argument, we provide an object of the class `list`, containing at least two elements, each of which contains the identifiers of the different lists to be compared.

For example, for the BP ontology at GO-Level 4, the $7(6)/2=21$ possible test calculated from the 7 gene lists contained in `allOncoGeneLists` are obtained as follows:
```{r, eval=FALSE}
eqTest_all_BP4 <- equivTestSorensen(allOncoGeneLists,
                                    geneUniverse = humanEntrezIDs, 
                                    orgPackg = "org.Hs.eg.db", 
                                    onto = "BP", 
                                    GOLevel = 4,
                                    d0 = 0.4444, 
                                    conf.level = 0.95)
```

But remember, it is much simpler if we already have the contingency tables as an object of the class `tableList.` In our case, we have already calculated the contingency tables for all possible pairs of `allOncoGeneLists` for the ontology BP, GO-Level 4, in Section 3.1.2 and stored them in the object `cont_all_BP4.` Therefore, we can calculate the `eqTest_all_BP4` object more efficiently in the following way:

```{r, eval=FALSE}
eqTest_all_BP4 <- equivTestSorensen(cont_all_BP4,
                                    d0 = 0.4444, 
                                    conf.level = 0.95)
```

```{r, echo=FALSE}
data(eqTest_all_BP4)
```

The result is an object of the class `equivSDhtestList`, which is exclusive to `goSorensen.` 

Remember that similar to the comparison of two lists in Section 4.1.1, you can use the function `upgrade` to update the results by changing the parameters of the tests, such as the confidence level, irrelevance limit, and others. For instance `upgrade(eqTest_atlas.sanger_BP4, d0 = 0.2857` to update the equivalence test using an irrelevance limit $d_0=0.2857$ 

Since the output contained in `eqTest_all_BP4` is very large, it is not displayed in this vignette. However, `goSorensen` provides accessor functions that allow you to retrieve specific outputs of interest. For example, to obtain a summary of the Sorensen dissimilarities contained in the tests comparing all pairs of lists in the BP ontology at GO-Level 4, you can use the function `getDissimilarity` and retrieve them as follows:
```{r, echo=FALSE}
options(digits = 4)
```


```{r}
getDissimilarity(eqTest_all_BP4, simplify = FALSE)
```

Another example of accessor function is the function `getPvalue` to obtain the p-values across the object `eqTest_all_BP4`.

```{r}
getPvalue(eqTest_all_BP4, simplify = FALSE)
```

NaN values occur when the test statistic cannot be calculated due to an indeterminacy. This happens when there is no joint enrichment between two lists (i.e., when $n_{11}=0$), indicating that it is impossible to determine equivalence between the compared lists.

In addition other accesor functions includes: `getSE` for the standard error, `getUpper` for the upper bound of the confidence interval and `getTable` for the enrichment contingency tables. 

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the object `eqTest_all_BP4`, which can be accessed using `data(eqTest_all_BP4)`.

Please consider this object only as illustrative example, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

## For More than One Ontology and GO-Level.
When you want to obtain the outputs of the equivalence tests to compare two or more lists across multiple ontologies and more than one GO-Level, you can use the function `allEquivTestSorensen`

For example to obtain the $7(6)/2=21$ equivalence tests calculated from the 7 gene lists in `allOncoGeneLists` for the three ontologies (BP, CC, and MF) and for the GO-Levels from 3 to 10, you can use the function `allEquivTestSorensen` as follows:
```{r, eval=FALSE}
allEqTests <- allEquivTestSorensen(allOncoGeneLists,
                                   geneUniverse = humanEntrezIDs, 
                                   orgPackg = "org.Hs.eg.db", 
                                   ontos = c("BP", "CC", "MF"), 
                                   GOLevels = 3:10,
                                   d0 = 0.4444, 
                                   conf.level = 0.95)
```

But remember, it is much simpler if we already have the contingency tables as an object of the class `allTableList` In our case, we have already calculated the contingency tables for all possible pairs of `allOncoGeneLists` for the ontologies BP, CC, MF, and for the GO-Levels 3 to 10, in Section 3.2 and stored them in the object `allContTabs` Therefore, we can calculate the `allEqTests` object more efficiently in the following way:
```{r, eval=FALSE}
allEqTests <- allEquivTestSorensen(allContTabs,
                                   d0 = 0.4444, 
                                   conf.level = 0.95)
```

The result is an object of the class `AllEquivSDhtest`, which is exclusive to `goSorensen.` 

Remember that similar to the equivalence tests in Section 4.1.1 and 4.1.2, you can use the function `upgrade` to update the results by changing the parameters of the tests, such as the confidence level, irrelevance limit, sample distribution (normal or bootstrap) and others.

Similarly to Sections 4.1.1 and 4.1.2, you can use the accessor functions to obtain a summary of the key outputs of the equivalence test, such as the Sorensen dissimilarities using `getDissimilarity`), p-values using `getPvalue`, enrichment contingency tables using `getTable`, and more.

#### NOTE: {.unnumbered}
To provide users with a quick visualization, the `goSorensen` package includes the object `allEqTests`, which can be accessed using `data(allEqTests)`.

Please consider this object only as illustrative example, valid exclusively for the `allOncoGeneLists` data contained in this package. Note that gene lists, GO terms, and Bioconductor may change over time. The current version of these results was generated with Bioconductor version 3.20.

# Bibliography. {.unnumbered}
Flores, P., Salicrú, M., Sánchez-Pla, A. et al. An equivalence test between features lists, based on the Sorensen--Dice index and the joint frequencies of GO term enrichment. BMC Bioinformatics 23, 207 (2022). <https://doi.org/10.1186/s12859-022-04739-2>


<!-- 
# Session information. {.unnumbered}

All software and respective versions used to produce this document are listed below.

```{r sessionInfo}
sessionInfo()
```
-->